%!TEX encoding = UTF-8 Unicode

\addchap[Acknowledgments]{Acknowledgments%
	%\foreignlanguage{french}{Remerciements} {\Large (Acknowledgments)}
}

\vspace*{-5mm}
First and foremost, I would like to express my sincere gratitude to my advisors, Nicolas \textsc{Gast} and Bruno \textsc{Gaujal}, for their excellent guidance and their attention toward me.
I have learnt a lot during my thesis thanks to them.
They taught me to be precised in my speech and to think clearly and independently.
Without them, I would not have come this far.

Secondly, I wish to thank Annie \textsc{Simon} and Arnaud \textsc{Legrand} for their empathy, their advice on some of my daily life problems in France, and their effort to facilitate the research life in Polaris team. I am grateful to Institut national de recherche en sciences et techonologies du numérique (Inria) for financing my thesis and Laboratoire d'Informatique de Grenoble (LIG) for hosting me.

I also want to thank the members of the jury for kindly accepting to be part of this committee.
In particular, thanks to the two reviewers, Aditya \textsc{Mahajan} and Konstantin \textsc{Avrachenkov}, for providing pertinent comments on my manuscript which helped to improve the final version.

My special thanks to my colleagues in Polaris and DataMove teams, to my officemates, Chen \textsc{Yan} and Till \textsc{Kletti}, for all nice conversations on various topics.
My deep thanks to my Cambodian seniors, Sotheara \textsc{Leang}, Rottana \textsc{Ly}, Sreynoch \textsc{Soung}, and juniors, Kimhong \textsc{Chao}, Bonpagna \textsc{Kann}, Pungpanhavoan \textsc{Tep}, Puthineath \textsc{Lay}, Sonita \textsc{Te}, Leangsiv \textsc{Han}, Sophal \textsc{Thear} for all fun moments during lunch and during our trips.
Thanks to all members of Association des Polytechniciens Khmers (AXK) for their encouragement and friendship.
I also want to thank Leangsiv \textsc{Han}, Mengchhoung \textsc{Ang}, and Nakanyseth \textsc{Vuth} for their helps before my departure to Cambodia.

I also enjoyed the friendship of several great people at LIG: Jibril \textsc{Frej}, Sebastian \textsc{Allmeier}, Louis-Sébastien \textsc{Rebuffi}, Victor \textsc{Boone}, Romain \textsc{Cravic}, and Fiorella \textsc{Albasini}.
I am glad that I have met you.

Last but not least, I am indebted to my parents, Sothea \textsc{Mong} and Sim \textsc{Veng}, for their belief in me, their endless support, and their unconditional love, to my brothers, Kim Khuy \textsc{Khun} and Kim Eng \textsc{Khun}, for their life experiences, their advice, and their helps, to my sisters, Sivlang \textsc{Khun} and Sivluy \textsc{Khun}, for their belief in me, their love, and their concern.

%\begin{otherlanguage}{french}
%
%Merci public !
%
%\end{otherlanguage}
